Towards typed repositories of proofs \\[0.6em] \small \textsf{MIPS 2010}

### 2010-07-07 talks/MIPS2010/slides.tex
Première passe sur l'explication de la syntaxe et des règles de typage
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@143 403e655e-980a-47bd-8c60-1ee5348bb6b7

Let me start with a simple observation relating the activity of both
the working mathematician and the computer scientist. Whereas both
activities are percieved, from the outside, as apodictic because the
final object of their study, the proof or the program, seems to be
unquestionably valid (or not).

@@

Here is a schematic view of the traditional way to manage formal
developments. The user writes scripts which are version controlled,
and these scripts are parsed and checked on demand. By pushing version
management inwards, we abstract from syntax, and more informations
become available. For instance we can compute the graph of
dependencies. At this level already, the situation is greatly
enhanced: syntax, and file-based development, get the place they
deserve: they are relegated to a tool for user interaction, the actual
development being stored and manipulated in an abstract form. We could
even imagine syntax being dynamically switchable, part of user
preferences!

@@

\begin{frame}{A core meta-language for incremental type-checking}
  \begin{columns}[t]
    \begin{column}{0.5\textwidth}
      \begin{block}{Expresses}
        \begin{itemize}
        \item (abstract) Syntax
        \item (object-) Logics
        \item Proofs (-terms)
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{block}{Features}
        \begin{itemize}
        \item Typing
        \item Incrementality
        \end{itemize}
      \end{block}
    \end{column}
  \end{columns}
  \vspace{2em}
  \begin{center}
    \emph{A kernel for a typed version control system?}
  \end{center}
\end{frame}

In the following, I will present a preliminary implementation of these
ideas. It is a meta-language for incremental proof-checking, that is a
language in which one can define abstract syntaxes, logical rules and
proofs, and allows to check validity, and express the incrementality
of the constructions. It can be viewed as a kernel for a generic,
typed version control system.

@@

\textsf{Git} is one of the most successful version control system, and
part of its success is due to the very simple, yet very efficient
storage model. It doesn't keep track of changes \emph{per se}, but of
similarities between version. Given a base directory structure,
already committed in its database, and a new version of it with, let's
say, one file changed -- \texttt{f3.txt}, it stores the new version by
linking all common subdirectory structure to the previous version.
This is possible because every node in the tree, directories and
files, have a unique name. So actually, the database of \textsf{git}
looks like this.

Moreover, similarities are easily tracked because every name is
carefully chosen to reflect the \emph{content} of a file or directory.
In practice, \texttt{git} uses a hash function to produce all names in
the database: files are given the hash of their content, and
directories are given the hash of the hash of all the files it
contains, recursively.

@@

\begin{frame}{Incremental type-checking}
  A small language for representing proofs:
  \only<1>{\[ t\ ::=\ (x:t)\cdot t\ |\ x\ |\ t\ t\ |\ *\] \[\]}%
  \only<2>{\[ t\ ::=\ (x:t)\cdot t\ |\ \alert x\ |\ \alert{t\ t}\ |\ *\] \[\]}%
  \only<3>{\[ t\ ::=\ (x:t)\cdot t\ |\ \alert a\ |\ *\]%
    \[ a\ ::=\ x\ |\ a\ x \]} 
  \only<4>{\[ t\ ::=\ (x:t)\cdot t\ |\ a\
    |\ *\ |\ \alert{(x = a)\cdot t}\]%
    \[ a\ ::=\ x\ |\ a\ x \]}
  \begin{example}
    \[ \infer[Lam]{\Gamma,A\vdash B}{\Gamma\vdash A\to B} \]
    \[ \textsc{Lam}\ :\ (\Gamma : env)\cdot(A B : type)\cdot
    (d : \Gamma,A\vdash B)\cdot \Gamma\vdash A\to B \]
  \end{example}
\end{frame}

>>>>

Let me start with a simple observation relating the activity of both
the working mathematician and the computer scientist. Both activities
could be percieved, from the outside, as apodictic because the final
object of their study, the proof or the program, seems to be
unquestionably valid (or not).

@@

Here is a schematic view of the traditional way to manage formal
developments. The user writes scripts which are version controlled,
and these scripts are parsed and checked on demand. By pushing version
management inwards, we abstract from syntax, and more informations
become available. For instance we can compute the acyclic graph of
dependencies. At this level already, the situation is greatly
enhanced: syntax, and file-based development, get the place they
deserve: they are relegated to a tool for user interaction, the actual
development being stored and manipulated in an abstract form. We could
even imagine syntax being dynamically switchable, part of user
preferences!

@@

\begin{frame}{A core meta-language for incremental type-checking}
  \begin{columns}[t]
    \begin{column}{0.5\textwidth}
      \begin{block}{Expresses}
        \begin{itemize}
        \item (abstract) Syntax
        \item (object-) Logics
        \item Proofs (-terms)
        \end{itemize}
      \end{block}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{block}{Features}
        \begin{itemize}
        \item Typing
        \item Incrementality
        \item Dependency
        \end{itemize}
      \end{block}
    \end{column}
  \end{columns}
  \vspace{2em}
  \begin{center}
    \emph{A kernel for a typed version control system?}
  \end{center}
\end{frame}

In the following, I will present a preliminary implementation of these
ideas. It is a meta-language for incremental proof-checking, that is a
language in which one can define abstract syntaxes, logical rules and
proofs, and allows to check validity, express the incrementality of
the constructions, and their dependency. It can be viewed as a kernel
for a generic, typed version control system.

@@

\textsf{Git} is one of the most successful version control system, and
part of its success is due to the very simple, yet very powerful
storage model. It doesn't keep track of changes \emph{per se}, but of
similarities between version. Given a base directory structure,
already committed in its database, and a new version of it with, let's
say, one file changed -- \texttt{f3.txt}, it stores the new version by
linking all common subdirectory structure to the previous version.
This is possible because every node in the tree, directories and
files, have a unique name. So actually, the database of \textsf{git}
looks like this: it associates to every name, a content (for files) or
a list of names (for directories). The invariant is that this database
forms a DAG. This naming policy makes all content adressable.

Moreover, similarities are easily tracked because every name is
carefully chosen to reflect the \emph{content} of a file or directory.
In practice, \textsf{git} uses a hash function to produce all names in
the database: files are given the hash of their content, and
directories are given the hash of the hash of all the files it
contains, recursively.

@@

\begin{frame}{A typed repository}
  % exemple
\end{frame}

on va plus loin grace a l'approche LF où tous les objets ont un type,
et font partie de l'arbre de syntaxe. Donc tout est hash-consé
(règles logiques, syntaxe, preuves).

\begin{frame}{Incremental type-checking, incrementally}%
  \begin{block}{Syntax}%
    \only<1>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ x\ |\ t\ t\ %
      |\ *\] \[\]}%
    \only<2>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert x\ |\ %
      \alert{t\ t}\ |\ *\] \[\]}%
    \only<3>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert a\ |\ %
      *\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<4>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ a\ |\ *\ |\ %
      \alert{(x = a)\cdot t}\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<5>{\[ t\ ::=\ \alert{[x:t]\cdot t}\ |\ (x:t)\cdot t\ |\ a\ %
      |\ *\ |\ (x = a)\cdot t\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
    \only<6->{\[ t\ ::=\ (x:t)\cdot t\ |\ a\ |\ *\ |\ (x = a)\cdot t\]%
      \[ a\ ::=\ x\ |\ a\ x \]}%
  \end{block}
  \begin{block}{Semantics}
      \[ \Gamma\ ::=\ \cdot\ |\ \Gamma [x:t]\only<4>{\ |\
        \alert{\Gamma[x=a:t]}}
      \only<5->{\ |\ \Gamma[x=a:t]}\] 
      \vspace{1em}
      \begin{center}
        \begin{overlayarea}{0.8\textwidth}{5em}
          \only<-6>{ $ \Gamma \vdash t : u $ \hfill `` In environment
            $\Gamma$, term $t$ has type $u$ '' } \only<7>{ $ \Gamma
            \vdash t : u \alert{\Rightarrow \Delta} $ \hfill ``
            From \alert{repository} $\Gamma$, term $t$ of type $u$ \alert{leads to
            the new repository $\Delta$}'' }
        \end{overlayarea}
      \end{center}
  \end{block}
\end{frame}

\begin{frame}{Typing rules}
  \begin{block}{Product - Init}
  \begin{mathpar}
    \large
    \infer                      % prod
    {\Gamma\vdash t : *\uncover<2->{ \Rightarrow \_} \and \Gamma[x:t]\vdash u :
      *\uncover<2->{\Rightarrow\Delta}}
    {\Gamma\vdash(x:t)\cdot u : *\uncover<2->{\Rightarrow\Delta}}
    \and
    \uncover<3->{                               %init
      \infer{ }
      {\Gamma\vdash x : t \uncover<4->{\Rightarrow\Gamma}}
      \quad {\small [x:t]\in\Gamma}
    }
  \end{mathpar}
  \end{block}
\end{frame}

\begin{frame}{Typing rules}
  \begin{block}{Equality binder}
    \begin{mathpar}
      \large
      \infer
      {\Gamma\vdash a : u \Rightarrow \_ \and \Gamma[x=a:u]\vdash t :
        *\Rightarrow\Delta }
      {\Gamma\vdash (x=a)\cdot t : * \Rightarrow\Delta}
      \quad {\small [y=a:u] \notin\Gamma}
      \and \pause
      \infer
      {\Gamma\vdash t \{y/x\} : * \Rightarrow\Delta}
      {\Gamma\vdash(x=a)\cdot t : * \Rightarrow\Delta}
      \quad {\small [y=a:u] \in\Gamma}
    \end{mathpar}
  \end{block}
  \pause
  \begin{block}{Application}
    \[ \infer
      {\Gamma\vdash a : (y:u)\cdot t \Rightarrow\Delta}
      {\Gamma\vdash a\ x : t \{x/y\} \Rightarrow\Delta}
      \quad [x : u] \in \Gamma
      \]
  \end{block}
\end{frame}

\begin{frame}{Naming}

\end{frame}

environnement : nom -> contenu * type

\begin{frame}{Conclusion}

\end{frame}
